---
title: "Practical 4: Panel Network Models"
author: "Developed by Philip Leifeld and delivered by Jingyi Kang"
date: "20 April 2026"
---

A preparatory step to avoid output getting sliced up into different boxes when compiling this document.

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

# Temporal ERGMs: A simple example with international alliances

## Data preparation

Load packages and data.
Load the `alliances` data, which contain bilateral defensive alliances among 164 countries over a 20-year span.
See the help page (`?alliances`) for more details about the data (after loading `btergm`).

```{r}
library("network")
library("ergm")
library("btergm")
library("texreg")

set.seed(12345)

data("alliances")
```

We will jump right in and estimate the first TERGM.
Note the linear time trend, autoregression term, and 50 bootstrap replications.
This will take a minute to complete because a pretty large logit model has to be estimated 50 times.
We should normally use something like `R = 1000` for effective bootstrapping, so keep in mind this low number is just for illustration in class.
We could speed this up massively by running this in parallel.
The help page for the btergm function (`?btergm`) contains details on how to do so.
At home, figure out how many cores you have in the computer you are working with, and try out parallel computation.

```{r}
m1 <- btergm(allyNet ~ edges + timecov(transform = function(t) t)
     + memory("autoregression") + gwesp(0, fixed = TRUE) + edgecov(LSP)
     + edgecov(warNet) + nodecov("polity") + nodecov("cinc") + absdiff("polity")
     + absdiff("cinc") + edgecov(contigMat), R = 50)
```

Display the output.

```{r}
summary(m1)
```

Think about the interpretation of the results for a moment.
E.g., what does each model term express substantively?
What is the difference between "Estimate" and "Boot mean"?
How is the confidence interval constructed, and what does it mean?

We can display the results with nicer labels in a proper table and think about it some more.
First, we need to create a mapping between the coefficient names in the results and the new names we want to use.

```{r}
map <- list("edges" = "Edges",
            "edgecov.timecov1[[i]]" = "Linear time",
            "edgecov.memory[[i]]" = "Autoregression",
            "edgecov.LSP[[i]]" = "Shared allies at $t-1$",
            "gwesp.fixed.0" = "GWESP",
            "edgecov.warNet[[i]]" = "War",
            "nodecov.polity" = "Polity score (sender)",
            "absdiff.polity" = "Polity score (absdiff)",
            "nodecov.cinc" = "Capabilities (sender)",
            "absdiff.cinc" = "Capabilities (absdiff)",
            "edgecov.contigMat[[i]]" = "Direct contiguity")
```

We can feed this map into `screenreg` (or `texreg`, `knitreg`, `htmlreg`, `plotreg`, or `wordreg`, depending out the desired output format).

```{r}
screenreg(m1, single.row = TRUE, custom.coef.map = map,
       custom.model.names = c("Bootstrapped MPLE"), include.aic = FALSE)
```

This was estimated using the `btergm` function, i.e., with MPLE and temporal bootstrapping.
We could do the same thing now with MCMC-MLE using the `mtergm` function.
This takes even longer (about an hour in this case, and it will often be prohibitively long, as in multiple human lifetimes) with complex and large network datasets or near-degenerate models.
We set `eval = FALSE` in the chunk options so this does not get evaluated by default.
If you want to run it, please do so at home, but the results are already shown on the slides.
```{r, eval = FALSE}
m2 <- mtergm(allyNet ~ edges + timecov(transform = function(t) t)
    + memory("autoregression") + gwesp(0, fixed = TRUE) + edgecov(LSP)
    + edgecov(warNet) + nodecov("polity") + nodecov("cinc")
    + absdiff("polity") + absdiff("cinc") + edgecov(contigMat))
```

Assess goodness of fit.
This will simulate from all 20 time points and put the GOF results for all time points into a single plot.
It takes a couple of minutes.
This can also be parallelised; see help pages.
We should normally do 1,000 simulations or more overall.
Since we have 19 time steps, 100 per time step would be plenty (i.e., the default setting `nsim = 100`).
Here, we do just 20 simulations per time step to keep the time it takes manageable for illustration.

```{r, out.width = "100%", fig.width = 8, fig.height = 5}
g <- gof(m1,
         nsim = 20,
         statistics = c(dsp, esp, deg, ideg, geodesic, walktrap.modularity))
plot(g)
```

# A more advanced TERGM example with node composition change

This example is about evolving friendships in a school class.
See `?knecht` for details.

```{r}
data("knecht", package = "btergm")
sapply(friendship, dim) # composition change over time...
```

Make sure the network matrices have row labels.
Otherwise, the function won't be able to match nodes between time points.
Here, all matrices initially have the same dimensions, and missing data are denoted using special numbers.
So, it's safe to name them as consecutive integer numbers.
We need to do the same thing with the primary school co-attendance covariate matrix and the sex node covariate.

```{r}
for (i in 1:length(friendship)) {
  rownames(friendship[[i]]) <- 1:nrow(friendship[[i]])
  colnames(friendship[[i]]) <- 1:ncol(friendship[[i]])
}
rownames(primary) <- rownames(friendship[[1]])
colnames(primary) <- colnames(friendship[[1]])
sex <- demographics$sex
names(sex) <- 1:length(sex)
```

Imputation of NAs and removal of absent nodes.

```{r}
friendship <- handleMissings(friendship, na = 10, method = "remove")
friendship <- handleMissings(friendship, na = NA, method = "fillmode")
```

Adjust dimensions and add nodal covariates to the networks.
Dimension adjustment means we shorten the `sex` vector where necessary by removing the entries that are missing in the friendship network at the respective time point after removing absent nodes in the previous step.
We then save it as a covariate in the network object at the respective time step.

```{r}
for (i in 1:length(friendship)) {
  s <- adjust(sex, friendship[[i]])
  friendship[[i]] <- network(friendship[[i]])
  friendship[[i]] <- set.vertex.attribute(friendship[[i]], "sex", s)
}
sapply(friendship, network.size) # check network dimensions over time
```

Plot the networks.

```{r, out.width = "100%"}
par(mfrow = c(2, 2), mar = c(0, 0, 1, 0))
for (i in 1:length(friendship)) {
  plot(network(friendship[[i]]), main = paste("t =", i),
  usearrows = TRUE, edge.col = "grey50")
}
```

For estimation, we could put the formula directly into the `btergm` or `mtergm` call as usual, but we'll save a few lines of code by first saving the formula and then re-using it a couple of times.
Reflect on what the different model terms mean.
Use the help pages and JStatSoft article on `btergm` to find out if you are unsure.

```{r}
f <- friendship ~ edges + mutual + delrecip + istar(2) + memory(type = "stability") +
    nodeofactor("sex") + nodeifactor("sex") + nodematch("sex") + edgecov(primary)
```

Use this formula in bootstrapped MPLE estimation.

```{r}
m3 <- btergm(f, R = 1000)
```

Estimate the same model using MCMC-MLE.
Note that we need to increase the sample size and MCMC interval to avoid degeneracy in sampling networks (and hence estimation).

```{r}
m4 <- mtergm(f, control = control.ergm(MCMC.samplesize = 5000, MCMC.interval = 2000))
```

Create a side-by-side regression table to compare results.

```{r}
map2 <- list("edges" = "Edges",
             "mutual" = "Reciprocity",
             "edgecov.delrecip" = "Delayed reciprocity",
             "edgecov.delrecip[[i]]" = "Delayed reciprocity",
             "istar2" = "Incoming two-stars",
             "edgecov.memory[[i]]" = "Edge stability",
             "edgecov.memory" = "Edge stability",
             "nodeofactor.sex.2" = "Sex = male (sender)",
             "nodeifactor.sex.2" = "Sex = male (receiver)",
             "nodematch.sex" = "Sex (node match)",
             "edgecov.primary[[i]]" = "Primary school co-attendance",
             "edgecov.primary" = "Primary school co-attendance")

screenreg(list(m3, m4), single.row = TRUE, custom.coef.map = map2,
       custom.model.names = c("btergm", "mtergm"))
```

Goodness of fit with `btergm`.

```{r, out.width = "100%", fig.width = 8, fig.height = 8}
plot(btergm::gof(m3, joint = FALSE))
```

Goodness of fit with `mtergm`.

```{r, out.width = "100%", fig.width = 8, fig.height = 8}
plot(btergm::gof(m4, joint = FALSE))
```

Out-of-sample prediction.
Estimate based on the first three time steps and predict the fourth, then run `gof` to see how well forward simulations match with the structure of the left-out fourth network.

```{r, out.width = "100%", fig.width = 8, fig.height = 3}
f5 <- friendship[1:3] ~ edges + mutual + delrecip + istar(2) +
    memory(type = "stability") + nodeofactor("sex") + nodeifactor("sex") +
    nodematch("sex") + edgecov(primary)
m5 <- btergm(f5, R = 1000)
g5 <- btergm::gof(m5, nsim = 1000, target = friendship[[4]],
    formula = friendship[3:4] ~ edges + mutual + delrecip + istar(2) +
    memory(type = "stability") + nodeofactor("sex") + nodeifactor("sex") +
    nodematch("sex") + edgecov(primary), coef = coef(m5),
    statistics = c(esp, triad.undirected, rocpr))
plot(g5, rgraph = TRUE)
```

We can also directly simulate new networks out of sample (here: two networks as an example).

```{r}
tp <- tergmprepare(f) # create objects taking into account composition change
s <- simulate_formula(tp$networks$t4 ~ edges + mutual + edgecov(tp$delrecip[[3]]) +
    istar(2) + edgecov(tp$memory[[3]]) + nodeofactor("sex") + nodeifactor("sex") +
    nodematch("sex") + edgecov(tp$primary[[3]]), coef = unname(coef(m5)),
    nsim = 2)
par(mfrow = c(1, 2), mar = c(0, 0, 1, 0))
lapply(s, plot)
```

# Exercise: Reparameterisation with `btergm`

Build on the alliance and school class friendship networks and examples by creating your own model specification and estimating it.
Use different model terms from the `ergm` package as usual, but also look at the help page for `"tergm-terms"`, and try out different temporal specifications, including `delrecip`, `memory`, and `timecov` with different plausible arguments.
Also vary the covariates you include.
Assess goodness of fit, and try out-of-sample prediction.