---
title: "Practical 3: ERGM: Estimation and Sampling"
author: "Developed by Philip Leifeld and delivered by Jingyi Kang"
date: "23 March 2026"
---

# ERGM estimation and diagnostics with the Swiss co-authorship dataset

The goal in this script is to replicate the analysis from Leifeld (2018) in Physica A: Statistical Mechanics (referenced on the lecture slides).

## Data preparation

Load packages and data.

```{r}
library("network")
library("sna")
library("ergm")
library("texreg")

seed <- 12345
set.seed(seed)

data("ch_coauthor", package = "btergm")
```

Create the `network` object, and store attributes.
```{r}
ch_nw <- network(ch_coaut, directed = FALSE)
set.vertex.attribute(ch_nw, "frequency", ch_nodeattr$num_publications)
set.vertex.attribute(ch_nw, "status", as.character(ch_nodeattr$status))
set.vertex.attribute(ch_nw, "male", ch_nodeattr$male)
set.vertex.attribute(ch_nw, "share_en_articles",
                     ch_nodeattr$share_en_articles)
```

Create a covariate matrix where a 1 indicates that two authors have the same affiliation.
We use some matrix multiplication to do this efficiently, but we could have used nested loops or other means.
```{r}
# create same affiliation matrix
ch_inst_indices <- which(grepl("^inst_.+", colnames(ch_nodeattr)))
ch_same_affiliation <- as.matrix(ch_nodeattr[, ch_inst_indices]) %*%
  t(ch_nodeattr[, ch_inst_indices])
```

Create a covariate matrix indicating whether two authors work at the same chair or in the same research group within a department.
To do so, we use nested loops and check if two authors have both the same chair title *and* the same institutional affiliation (because the same chair title could exist in two universities, for example "Comparative Politics").
```{r}
# create same chair matrix
ch_nodeattr$chairtitle[ch_nodeattr$chairtitle == ""] <- NA
ch_same_chair <- matrix(0, nrow = nrow(ch_same_affiliation),
                        ncol = ncol(ch_same_affiliation))
for (i in 1:length(ch_nodeattr$chairtitle)) {
  for (j in 1:length(ch_nodeattr$chairtitle)) {
    if (i != j &&
        !is.na(ch_nodeattr$chairtitle[i]) &&
        !is.na(ch_nodeattr$chairtitle[j]) &&
        ch_nodeattr$chairtitle[i] == ch_nodeattr$chairtitle[j] &&
        ch_same_affiliation[i, j] == TRUE) {
      ch_same_chair[i, j] <- 1
    }
  }
}
rownames(ch_same_chair) <- rownames(ch_same_affiliation)
colnames(ch_same_chair) <- colnames(ch_same_affiliation)
```

Create a supervision covariate matrix using element-wise multiplication of the same-affiliation matrix, same-chair matrix, and a new matrix indicating that one author is a professor while the other one isn't.
```{r}
# create supervision matrix (same chair + affiliation + mixed seniority)
ch_supervision <- ch_same_affiliation *
  ch_same_chair *
  matrix(ch_nodeattr$status == "professor",
         nrow = nrow(ch_same_chair),
         ncol = ncol(ch_same_chair),
         byrow = FALSE) *
  matrix(ch_nodeattr$status != "professor",
         nrow = nrow(ch_same_chair),
         ncol = ncol(ch_same_chair),
         byrow = TRUE)
```

## ERGM estimation

We are now ready to estimate the full ERGM reported in Leifeld (2018) in Physica A: Statistical Mechanics and its Applications (referenced on the lecture slides).
We use the `control` argument to supply additional MCMC-MLE tweaks for the estimation; don't worry about them for now, but do read up on `control.ergm` in the help pages next week.
At home, look at the paper and make sure you understand what's happening in this model and how to interpret the numbers.
```{r}
ch_model <- ergm(ch_nw ~
                   edges +
                   gwesp(0.3, fixed = TRUE) +
                   gwdegree(0.4, fixed = TRUE) +
                   nodecov("frequency") +
                   nodefactor("status") +
                   nodefactor("male") +
                   nodematch("male") +
                   edgecov(ch_dist100km) +
                   edgecov(ch_same_affiliation) +
                   edgecov(ch_same_chair) +
                   edgecov(ch_supervision) +
                   edgecov(ch_topicsim) +
                   nodecov("share_en_articles") +
                   edgecov(ch_en_article_sim),
                 control = control.ergm(MCMLE.termination = "Hummel",
                                        MCMLE.effectiveSize = NULL))
summary(ch_model)  # corresponds Column 1 in Table 3 in Leifeld (2018)

# regression table using the texreg package
screenreg(list(ch_model), single.row = TRUE)
```

## Goodness of fit

Assess goodness of fit.
```{r}
ch_gof <- gof(ch_model)
plot(ch_gof)  # visual inspection of GOF
```

## Model convergence and degeneracy evaluation

MCMC diagnostics.
```{r}
mcmc.diagnostics(ch_model)
```

# Exercise: Simulation
Simulate new networks from the estimated model above.
Do the simulations look like the original network you have visualised and explored so often?
Can you come up with other metrics or criteria to assess the goodness of fit by comparing simulated with observed networks than the ones implemented in the `gof` function, perhaps by taking into account the exogenous covariates or other criteria?

# Exercise: Play with the model specification
Reparameterise the model with different specifications.
Compare the model fit and MCMC diagnostics.
Can you come up with a specification that also fits well but comes to different substantive conclusions?

# Exercise: Application to the Putin dataset
Estimate a bipartite (two-mode) ERGM for the Putin dataset.
Select the right model terms that work well and make sense in a bipartite setting.
Can you find a specification that works well with the data?
Why could it be a bad idea to apply ERGMs to this dataset?
