---
title: "Practical 6: Permutation Approaches and Latent Space Models"
author: "Developed by Philip Leifeld and delivered by Jingyi Kang"
date: "11 May 2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

# Conditional uniform graph test (CUG)

## Load packages and data

Install the `statnet.data` package if necessary.

```{r}
packageVersion("statnet.data") # check if package is installed
# library("remotes")
# install_github("statnet/statnet.data")
```

Load packages and data.
```{r}
library("network")
library("sna") # tested with version 2.8
library("ergm")
library("tidygraph")
library("ggraph")
library("statnet.data")
library("vegan")
library("lsa")
library("texreg")

data("chemnet", package = "btergm")
```

## Manual CUG test

Plot reciprocal edges in a different colour.
```{r}
plot(network(pol),
     edge.col = 1 * (pol == 1 & t(pol) == 1) + 1,
     vertex.col = "green")
```

Derive the parameters for the CUG test.
```{r}
n <- nrow(pol) # number of nodes
e <- sum(pol) # number of edges (fixed in the simulations)
non_e <- n * n - n - e # subtract diagonal and edges
observed <- sum(pol == 1 & t(pol) == 1) / 2
observed # observed reciprocity
```

Save matrix indices for valid dyads (i.e., off-diagonal).
```{r}
offdiag_indices <- which(row(pol) != col(pol))
head(offdiag_indices, 40) # show first 20 results; note how 1 and 32 are missing
```

Create a vector with the right number of edges and non-edges.
```{r}
dyads <- c(rep(1, e), rep(0, non_e))
```

Create an empty matrix.
```{r}
mat <- matrix(0, nrow(pol), ncol(pol))
```

Overwrite the off-diagonal values in the matrix with the zeros and ones in a random (i.e., resampled/permuted/shuffled) order, and calculate reciprocity.
Repeat 10,000 times.
```{r}
results <- sapply(1:10000, function(x) {
  mat[sample(offdiag_indices)] <- dyads
  sum(mat == 1 & t(mat) == 1) / 2
})
```

Locate the observed reciprocity value in the CUG-based null distribution.
Is it in the tails?
I.e., is it unexpectedly large or small?
```{r}
xlim <- c(min(c(results, observed)) - 10,
          max(c(results, observed)) + 20)
hist(results, xlim = xlim, xlab = "")
abline(v = quantile(results, c(0.025, 0.975)),
       col = "blue", lwd = 3, lty = 1)
abline(v = observed, col = "red", lwd = 3)
```

## CUG test using the `cug.test` function

You can get various network statistics, such as reciprocity, using the `summary` method in the `ergm` package.
To do so, write a formula modelling the network as a function of reciprocity (`mutual`) and other summary statistics added on the right-hand side of the tilde (`~`).
Check out the help pages for `?"ergm-terms"` for a list of terms you can add.
```{r}
# same with the ergm package
summary(network(pol) ~ mutual + edges + istar(2) + twopath)
```

Another example with two-paths.
This time, we won't do it manually but use the `cugtest` function in the `sna` package.
First, create a function that calculates two-paths.
Then, apply it both to the observed network and permuted networks to create a reference distribution.
```{r}
f <- function(x) summary(x ~ twopath)
obs_twopath <- f(pol)
results <- cug.test(dat = pol,
                    FUN = f,
                    mode = "digraph",
                    cmode = "edges")
plot(results)

# p-value:
1 - ecdf(results$rep.stat)(obs_twopath)
```

# Exercise: Practice CUG

Apply what you have learned about CUG tests to the other empirical networks you worked with in previous sessions, including `chemnet`, the co-authorship network, Rob Cross's organisational network, and Putin's network.

Explore which network statistics from the ergm package (see `?"ergm-terms"`) could be useful to describe the network structure, and test their univariate effects using CUG tests.

# Plotting multigraphs (recap)

```{r}
mar <- as.matrix(flomarriage)
bus <- as.matrix(flobusiness)
g_mar <- as_tbl_graph(mar, directed = FALSE) |>
  activate("edges") |>
  mutate(type = "marriage")
g_bus <- as_tbl_graph(bus, directed = FALSE) |>
  activate("edges") |>
  mutate(type = "business")

g_combined <- graph_join(g_mar, g_bus) |>
  activate("nodes") |>
  mutate(wealth = get.vertex.attribute(
    flomarriage, "wealth"))

# look at the data structure
print(g_combined, n = 2)
```

# Quadratic assignment procedure (QAP)

## Distance measures

```{r}
N <- matrix(c(0, 1, 0, 1, 1,
              0, 0, 0, 1, 1,
              0, 1, 0, 0, 1,
              0, 0, 1, 0, 0,
              0, 1, 0, 1, 0),
            byrow = TRUE, nrow = 5)

M <- matrix(c(0, 1, 0, 0, 1,
              1, 0, 0, 1, 0,
              0, 1, 0, 0, 1,
              1, 0, 1, 0, 0,
              0, 1, 0, 0, 0),
            byrow = TRUE, nrow = 5)

mat <- matrix(c(N, M),
              nrow = 2,
              byrow = TRUE)
mat

# Jaccard distance (for binary data)
vegdist(mat, method = "jaccard")[1] # vegan package

# Euclidean distance (for weighted data)
dist(mat, method = "euclidean")[1]

# correlation (for weighted data)
cor(mat[1, ], mat[2, ])

# cosine similarity (for weighted data)
cosine(mat[1, ], mat[2, ])[1, 1] # lsa package
```

## Manual implementation of bivariate QAP:

```{r}
# Jaccard similarity (i.e., 1 - distance)
jac <- 1 - vegdist(matrix(c(mar, bus), nrow = 2, byrow = TRUE), method = "jaccard")
jac
n <- nrow(mar)
qap <- sapply(1:10000, function(x){
  s <- sample(n) # new order to rearrange marriage matrix
  mar_perm <- mar[s, s] # apply new order to rows and columns
  1 - vegdist(matrix(c(mar_perm, bus), nrow = 2, byrow = TRUE), method = "jaccard")
})
hist(qap, main = "", xlab = "", xlim = c(0, jac))
abline(v = quantile(qap, c(0.025, 0.975)), col = "blue", lwd = 3, lty = 1)
abline(v = jac, col = "red", lwd = 3)
1 - ecdf(qap)(jac) # QAP p-value
```

## Implementation in the `sna` package

The `vegdist` function is a bit slow and not very convenient to deal with.
It does not accept two matrices but tries to apply the Jaccard coefficient to pairs of rows in a single matrix.
We will therefore write our own implementation using a list of two matrices.
```{r}
f_jac <- function(a, b) { # a and b are two adjacency matrices
  only_a <- sum(a * (1 - b))
  only_b <- sum((1 - a) * b)
  a_and_b <- sum(a * b)
  denominator <- a_and_b + only_a + only_b
  if (denominator == 0) { # prevent division by zero
    return(1) # minimal dissimilarity if everything is zero
  } else {
    return(a_and_b / denominator)
  }
}
```

Try out the new function and observe it gives us the same similarity as reported above.
```{r}
f_jac(bus, mar)
```

We can now use it in the `qaptest` function in the `sna` package.

```{r}
q <- qaptest(dat = bus,
             FUN = f_jac,
             b = mar,
             reps = 10000)
summary(q)
plot(q)
```

## Multiple regression quadratic assignment procedure (MRQAP)

### Linear model QAP (for weighted networks)

The examples here are taken from the `netlm` help page.
First, we create a data structure of four random graphs to try it out with.
These become our independent variables.
The `rgraph` function stacks four random graphs into a three-dimensional array, where the first index is for the network and the second and third indices refer to the row and column.
```{r}
x <- rgraph(20, 4) # 4 adjacency matrices with 20 nodes
```

Next, we generate an adjacency matrix with a linear combination of them with a known structure of associations.
This becomes our dependent variable.
Not that the fourth network remains unrelated to `y`.
```{r}
y <- x[1, , ] + 4 * x[2, , ] + 2 * x[3, , ]
```

Finally, fit a linear model with QAP test to recover the coefficients.
We'll try the original QAP version where the dependent network is permuted.
```{r}
nl1 <- netlm(y, x, nullhyp = "qapy", reps = 1000)
summary(nl1)
```

We can also try Dekker's "semi-partialling plus" procedure, which fixes some biases.
They yield very similar results here because the networks are random graphs.
```{r}
nl2 <- netlm(y, x, nullhyp = "qapspp", reps = 1000)
summary(nl2)
```

### Logit model QAP (for binary networks)

If we have a binary dependent network, we need to use a logit model in the QAP test.
This is implemented in the `netlogit` function.
First, we create the dependent data structure again via simulation.
We follow a logit model, i.e., we put the linear combination from above through an inverse logit link function and then sample based on the resulting probabilities to create a binary dependent network.
```{r}
y.p <- apply(y, c(1, 2), function(a) {
  1 / (1 + exp(-a))
})
y <- rgraph(20, tprob = y.p)
```

We can then apply the `netlogit` function in a similar way as the `netlm` function above.
```{r}
nl3 <- netlogit(y, x, nullhyp = "qapy", reps = 1000)
summary(nl3)
nl4 <- netlogit(y, x, nullhyp = "qapspp", reps = 1000)
summary(nl4)
```

These models are supported by the `texreg` package.
Create a text-based regression table:
```{r}
screenreg(list(nl3, nl4))
```

Or create an HTML regression table:
```{r, results = "asis"}
knitreg(list(nl3, nl4), caption = "Models 3 and 4 with netlogit.")
```

# Exercise: Practice bivariate QAP

Apply what you have learned about QAP tests to the other empirical networks you worked with in previous sessions, including `chemnet`, the co-authorship network, and Rob Cross's organisational network (by testing for an association with a departmental co-membership matrix).

# Exercise: Practice MRQAP

Apply what you have learned about MRQAP to the other empirical networks you worked with in previous sessions, including `chemnet`, the co-authorship network, and Rob Cross's organisational network (by testing for an association with a departmental co-membership matrix).

# Latent Space Models

## Latent space models using `latentnet`

Load the `latentnet` package and prepare the data.
We want to explain political and strategic information exchange as the dependent variable.
One of the explanatory variables we want to use is whether a an actor is a government actor.
I.e., do government actors receive more incoming ties (e.g., because other actors want to lobby them)?
We need to accommodate a binary government variable in the network object as an attribute for this purpose.

```{r}
library("latentnet")
data("chemnet", package = "btergm")
set.seed(12345)
pol_nw <- network(pol, directed = TRUE)
set.vertex.attribute(pol_nw, "gov", types$type == "gov")
```

Estimate a Euclidean latent space model with two dimensions and a couple of covariates.

```{r}
m1 <- ergmm(pol_nw ~ euclidean(d = 2) + edgecov(infrep) + nodeifactor("gov"))
```

Estimate a bilinear latent space model with two dimensions and sender random effect.
Increase the MCMC burn-in period and interval for better convergence.
This makes it take a moment longer, but it's necessary for making the estimation converge in this case.

```{r}
m2 <- ergmm(pol_nw ~ bilinear(d = 2) + rsender + edgecov(infrep) + nodeifactor("gov"),
            control = ergmm.control(burnin = 50000, interval = 100))
```

Estimate a (Euclidean) latent cluster model with four clusters.

```{r}
m3 <- ergmm(pol_nw ~ euclidean(d = 2, G = 4) + edgecov(infrep) + nodeifactor("gov"),
            control = ergmm.control(burnin = 30000, interval = 50))
```

Plot the models.

```{r, out.width = "100%", fig.width = 9, fig.height = 3, warning=FALSE}
par(mfrow = c(1, 3))
plot(m1)
plot(m2)
plot(m3)
```

Show the regression table with all results.

```{r}
screenreg(list(m1, m2, m3))
```

## Latent factor models with `amen`

Load package and data.

```{r}
library("amen")
data("chemnet", package = "btergm")
set.seed(12345)
```

Prepare data structures.
`amen` accepts a 3D array of dyadic covariates via the `Xdyad` argument, the equivalent of `statnet`'s `edgecov` term, but with multiple matrices stacked.
It also accepts separate matrices `Xcol` and`Xrow` with column vectors representing incoming receiver (and outgoing sender, respectively) node covariates.
We supply influence attribution via `Xdyad` and a government actor type dummy variable via `Xcol`.

```{r}
edgecovs <- array(numeric(), dim = c(nrow(pol), ncol(pol), 1)) # 30 x 30 x 1
edgecovs[, , 1] <- infrep
nodeifactors <- matrix(NA, nrow = nrow(pol), ncol = 1)
nodeifactors[, 1] <- 1 * (types$type == "gov")
```

We can also supply row (`rvar`), column (`cvar`), dyad (`dcor`), or node (`nvar`, in undirected networks) random effects (i.e., additive effects).
Here, we have a directed network, so in Model `m5`, we use the first three to control for additional variation that would otherwise be captured by network statistics in an ERGM.
We can also supply two-dimensional multiplicative effects, i.e., a bilinear latent space component, via `R = 2`.
`pol` is binary, so we need a logit link function.
This is achieved by setting `family = "bin"`.
The `burn` and `nscan` arguments determine the number of MCMC iterations (10,000 by default) and burn-in iterations that are discarded (500 by default).
We'll use the defaults here.
This would need to be increased in more complex models, e.g., for larger networks.
See `?ame` for additional arguments and options.

```{r}
# Bilinear (= multiplicative) model with two dimensions without random effects
m4 <- ame(Y = pol, Xdyad = edgecovs, Xcol = nodeifactors, family = "bin", R = 2,
          rvar = FALSE, cvar = FALSE, dcor = FALSE, nvar = FALSE, plot = FALSE)

# with additive component: random effects for rows, columns, dyads (but not nodes)
m5 <- ame(Y = pol, Xdyad = edgecovs, Xcol = nodeifactors, family = "bin", R = 2,
          rvar = TRUE, cvar = TRUE, dcor = TRUE, nvar = FALSE, plot = FALSE)
```

MCMC diagnostics and goodness of fit.

```{r, out.width = "100%", fig.width = 9, fig.height = 7}
plot(m5)
```

Plot the network in the estimated latent bilinear space.

```{r, out.width = "100%", fig.width = 9, fig.height = 7}
netplot(Y = pol, X = m5$UVPM, plotnames = TRUE)
```

For comparison, the network without the latent space, just directly plotted using the common Fruchterman-Reingold layout algorithm.

```{r, out.width = "100%", fig.width = 9, fig.height = 7}
netplot(Y = pol, X = xnet(pol), plotnames = TRUE)
```

# Exercise: Latent space model applications
Estimate latent space models with `latentnet` and `amen` for Rob Cross's organisational dataset.
Reflect on your modelling choices regarding space, dimensionality, covariates, random effects, and clustering.
Compare your results to the ERGM you estimated before.
Do you see any differences?

When you are done, conduct such an analysis also for the Swiss political science co-authorship network dataset, and follow the same instructions as noted above.